---
title: "Week 2 Discussion"
author: "Shaun Harrington"
format: html
editor_options: 
  chunk_output_type: console
---

## Setup

```{r}
#| warning: false

knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)

library(tidyverse)
library(fredr)
library(fpp3)

theme_set(theme_bw())

```

## Preliminary Analysis

### Get PCE Inflation Data

```{r}
 
data <- fredr_series_observations("PCEPI") %>% 
  mutate(date = yearmonth(date)) %>% 
  tsibble(index = date, key = series_id) %>% 
  mutate(
    diff.yoy = value / lag(value, n = 12) - 1,
    diff.mom = value / lag(value, n = 1) - 1
  )

train <- data %>% 
  filter(year(date) < 2022)
test <- data %>% 
  filter(year(date) >= 2022)

```

### Data Exploration

#### Time Plots 

```{r}
#| fig-height: 4

data %>% 
  autoplot(value) +
  ggtitle("Personal Consumption Expenditures: Chain-type Price Index")

```

```{r}
#| fig-height: 4

data %>% 
  autoplot(diff.yoy) +
  ggtitle("Year-Over-Year Percent Change")

```

#### Seasonality Plots

No apparent seasonality. 

```{r}
#| fig-height: 4

data %>% 
  gg_season(diff.mom) +
  ggtitle("Year-Over-Year Percent Change")

```


## Modeling

#### ETS Models

I'm using an additive Holt model, a multiplicative Holt model, and a damped multiplicative Holt model. My approach was a bit of trial and error here but I found a phi = .975 to be a reasonable inflation expectation. 

```{r}

# Guerrero lambda
lambda <- train %>% 
  features(value, features = guerrero) %>% 
  pull(lambda_guerrero)

fit <- train %>% 
  model(
    "holt.add" = ETS(box_cox(value, lambda) ~ error("A") + trend("A") + season("N")),
    # "hw.add" = ETS(box_cox(value, lambda) ~ error("A") + trend("A") + season("A")),
    "holt.mult" = ETS(box_cox(value, lambda) ~ error("M") + trend("A") + season("N")),
    # "hw.add.damp" = ETS(box_cox(value, lambda) ~ error("A") + trend("Ad") + season("N")),
    # "hw.mult.damp.98" = ETS(box_cox(value, lambda) ~ error("M") + trend("Ad", phi = .98) + season("N")),
    # "hw.mult.damp.99" = ETS(box_cox(value, lambda) ~ error("M") + trend("Ad", phi = .99) + season("N")),
    "hw.mult.damp.975" = ETS(box_cox(value, lambda) ~ error("M") + trend("Ad", phi = .975) + season("N"))
  )


```

The two multiplicative methods minimize the AICc. 

```{r}
glance(fit)
```


### Forecast

The models were trained on data prior to 2022, just before inflation levels peaked which will make for a good model comparison. The forecast period is 2022-2024. I extended this beyond today because we should expect inflation levels to converge to 2% in the long-run.

The damped model does perform significantly better than the other two models.

```{r}

  fx <- fit %>% 
        forecast(h = 36)
  

  fx %>% 
    autoplot(
        test,
        level = NULL
      ) +
        ggtitle("PCE Price Index Forecast")

```

Calculating the year-over-year inflation rate gives the following chart. 

```{r}

  fx.diff <- full_join(
    data %>% 
      select(date, value) %>% 
      rename(actual = value) %>% 
      as_tibble(),
    fx %>% 
      select(.model, value, .mean) %>% 
      hilo(level = c(85)) %>% 
      select(.model, .mean) %>% 
      pivot_wider(names_from = .model, values_from = .mean) %>% 
      as_tibble(),
    by = "date"
  ) %>% 
    arrange(date) %>% 
    mutate(
      across(-c(date, actual), \(x){
        ifelse(year(date) >= 2021, coalesce(x, actual), NA)
      }),
      across(-date, \(x){ifelse(year(date) >= 2021,x / lag(x, n = 12) - 1, NA)}),
      date = as.Date(date)
    ) %>%
    filter(date >= today() - years(2), date <= today() + years(10)) %>% 
    pivot_longer(-date, names_to = "forecast") %>% 
    filter(!is.na(value))

  fx.diff %>% 
    ggplot(aes(x = date, y = value, color = forecast, linetype = forecast)) +
    geom_line(size = 1) +
    geom_vline(xintercept = ymd("2022-1-01"), linetype = "dashed") +
    scale_y_continuous(labels = scales::label_percent()) +
    scale_color_manual(values = c("black", viridis::plasma(5))) +
    scale_linetype_manual(values = c("dashed", rep("solid", 5))) +
    ggtitle("PCE Year-Over-Year Forecast")

```

Though significantly better than the non-damped approaches, the damped model approaches 1% in the long run which is lower than we'd expect. There is a trade-off with the phi parameter which adjusts this long-term convergence. As phi approaches 1, the damped model approaches the forecasts of the non-damped, providing a worse forecast in the short-term. The selection of .975 aims to provide value in the short-term while not providing a bad forecast in the long-run, though it must be taken with some skepticism. 

The problem with the non-damped models is that they assume a constant trend across time. Exogenous factors play a role in the determination of inflation rates. With the Federal Reserve having raised rates to suppress inflation, this action would invalidate the constant trend assumption given in the non-damped models.